---
title: "ADA Final Project"
author: "Fariha Chowdhury"
date: "2024-12-11"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Install and load the packages
```{r}
library(tidyverse)
library(dplyr)
library(tidyr)
```



##loading the dataset
```{r}
data <- read.csv("/Users/farihaislamchowdhury/Desktop/Capstone/Fariha Capstone dataset.csv")
```

# View the first few rows of the dataset
```{r}
head(data)
```


# Check the structure of the dataset to understand variable types
```{r}
str(data)
```


# Summarize the dataset to get an overview of variable distributions
```{r}
summary(data)
```


### Check for missing values in each column
```{r}
colSums(is.na(data))
```


##separate dataset
```{r}
# Create a new dataset with the selected variables
data_new <- data %>%
 select(CHCCOPD3, EXERANY2, X_INCOMG1, X_AGE65YR, X_RFSMOK3)

view(data_new)
```

##Recode COPD Diagnosis
```{r}
data_new <- data_new %>%
  mutate(CHCCOPD3 = case_when(
    CHCCOPD3 == 1 ~ 1,  # Yes becomes 1
    CHCCOPD3 == 2 ~ 0,  # No becomes 0
    TRUE ~ NA_real_      # Missing or invalid values become NA
  ))
```



### Recode Physical Activity
```{r}
data_new <- data_new %>%
  mutate(EXERANY2 = case_when(
    EXERANY2 == 1 ~ "Yes",
    EXERANY2 == 2 ~ "No",
    TRUE ~ NA_character_
  ))
```

# Recode Smoking Status
```{r}
data_new <- data_new %>%
  mutate(X_RFSMOK3 = case_when(
    X_RFSMOK3 == 1 ~ "Former Smoker",
    X_RFSMOK3 == 2 ~ "Current Smoker",
    TRUE ~ NA_character_
  ))
```








# Recode _INCOMG1 variable##remember there is one new variable income category now which we will use
```{r}
data_new <- data_new %>%
  mutate(X_INCOMG1 = case_when(
    X_INCOMG1 %in% c(1, 2, 3, 4) ~ "<$50,000",  # Assign label for ≤$50,000
    X_INCOMG1 %in% c(5, 6, 7) ~ ">$50,000",    # Assign label for >$50,000
    X_INCOMG1 == 9 ~ NA_character_             # Assign NA for missing
  ),
  # Convert to factor with ordered levels
  X_INCOMG1 = factor(X_INCOMG1, levels = c("<$50,000", ">$50,000"))
  )
```







### Recode Gender(OMIT)--should consider as one of the limitation why we didnt chose gender since there are lots of NA's
```{r}
##data_new <- data %>%
  mutate(BIRTHSEX = case_when(
    BIRTHSEX == 1 ~ "Male",
    BIRTHSEX == 2 ~ "Female",
    TRUE ~ NA_character_
  ))
```






### Recode _AGE65YR variable##remember there is one new variable age category now which we will use
```{r}
data_new <- data_new %>%
  mutate(X_AGE65YR = case_when(
    X_AGE65YR == 1 ~ 0,              # Ages 18-64 remain as 0
    X_AGE65YR == 2 ~ 1,              # Ages 65 remain as 1
    X_AGE65YR == 3 ~ NA_real_        # Missing/Refused/Don't know becomes NA
  ),
  # Convert to factor with meaningful labels
  X_AGE65YR = factor(X_AGE65YR, levels = c(0, 1), labels = c("18-64", "65 or older"))
  )

# Verify the recoding
table(data_new$X_AGE65YR, useNA = "ifany")
```


##checking for missing variables
```{r}
data_new <- na.omit(data_new)
```




###Renaming the variable names
```{r}
# Rename variables in data_new
data_new <- data_new %>%
  rename(
    COPD_Diagnosis = CHCCOPD3,     # Rename CHCCOPD3 to COPD_Diagnosis
    Physical_Activity = EXERANY2,  # Rename EXERANY2 to Physical_Activity
    Income_Category = X_INCOMG1,   # Rename X_INCOMG1 to Income_Category
    Age_Category = X_AGE65YR,      # Rename X_AGE65YR to Age_Category
    Smoking_Status = X_RFSMOK3     # Rename X_RFSMOK3 to Smoking_Status
  )

# View the dataset with renamed variables
head(data_new)
```

###check the column names
```{r}
colnames(data_new)
```

```{r}
write.csv(data_new, "ADA_CLEANED_DATASET.csv", row.names = FALSE)
```



#####figure for BRFSS
```{r}
library(DiagrammeR)

grViz("
digraph flowchart {
  
  # Global node style
  node [fontname = Helvetica, shape = rectangle, fontsize=10]
  
  # Define nodes
  node1 [label = \"@@1\"]
  node2 [label = \"@@2\"]
  node3 [label = \"@@3\"]
  node4 [label = \"@@4\"]
  node5 [label = \"@@5\"]
  node6 [label = \"@@6\"]
  node7 [label = \"@@75\"]
  
  # Define edges
  node1 -> node2 -> node3 -> node4 -> node5 -> node6 -> node7 
}

[1]: \"Original BRFSS dataset:\\nN = 433,323\"
[2]: \"Randomly selected first 10,000 observations:\\nN = 10,000\"
[3]: \"Recoded Variables:\\n- COPD Diagnosis: 0/1\\n- Physical Activity: Yes/No\\n- Income Category: < $50,000 or > $50,000\\n- Age: <65 or ≥65\\n- Smoking status: Former Smoker/Current Smoker\"
[4]: \"Excluded invalid responses from all 5 variables (Refused, Not Sure)\"
[5]: \"Excluded missing values from these 5 variables\"
[6]: \"Total Excluded 2341 observations\"
[7]: \"Final analytic dataset:\\nN = 7,659\"
")
```





###table1 and other packages
```{r}
pacman::p_load(tidyverse, readr, table1, DiagrammeR, rsvg)
```

###table1
```{r}
table1(~ Physical_Activity + Income_Category + Age_Category + Smoking_Status | COPD_Diagnosis, data = data_new)
```




##Install and load package
```{r}
# Load the car package
library(car)
```

##Assumption 1: Multicolinearity
```{r}
model_vif <- glm(COPD_Diagnosis ~ Physical_Activity + Income_Category + Age_Category + Smoking_Status,family = "binomial", 
                data = data_new)
vif_values <- vif(model_vif)
print(vif_values)
```


##Assumption 2: Binary Dependent Variable
```{r}
table(data_new$COPD_Diagnosis)
str(data_new$COPD_Diagnosis)
```

##Assumption:3 Hosmer-Lemeshow test to see model fit
```{r}
install.packages("ResourceSelection") # Run this if not already installed
library(ResourceSelection)
hoslem_test <- hoslem.test(data_new$COPD_Diagnosis, fitted(model))
print(hoslem_test)
```

Interpretation: The logistic regression model appears to have a reasonably good fit to the data since the p-value is close to but still above 0.05.
However, the result is borderline, so it may warrant further investigation (e.g., additional diagnostics or recalibration of the model).


###Model for analysis(Logistic regression)
```{r}
# Fit the logistic regression model
model <- glm(COPD_Diagnosis ~ Physical_Activity + Income_Category + Age_Category + Smoking_Status,
             family = binomial, data = data_new)

# Extract Odds Ratios, Confidence Intervals, and P-values
odds_ratios <- exp(coef(model))  # Calculate Odds Ratios
conf_int <- exp(confint(model))  # Calculate Confidence Intervals
p_values <- coef(summary(model))[, "Pr(>|z|)"]  # Extract p-values

# Format p-values to 3 decimal points, replacing very small values with "<0.001"
p_values_rounded <- ifelse(p_values < 0.001, "<0.001", sprintf("%.3f", p_values))

# Combine results into a table
results <- data.frame(
  Variable = rownames(coef(summary(model))),
  Odds_Ratio = round(odds_ratios, 3),
  CI_Lower = round(conf_int[, 1], 3),
  CI_Upper = round(conf_int[, 2], 3),
  P_Value = p_values_rounded
)

# Print the results table
print(results)

# Save the results table as a CSV file (optional)
write.csv(results, "Logistic_Regression_Results.csv", row.names = FALSE)

# Optional Visualization: Forest plot of odds ratios with confidence intervals
library(ggplot2)
results$Variable <- factor(results$Variable, levels = rev(results$Variable))  # Reverse the order for better plotting
ggplot(results, aes(x = Odds_Ratio, y = Variable)) +
  geom_point() +
  geom_errorbarh(aes(xmin = CI_Lower, xmax = CI_Upper), height = 0.2) +
  labs(title = "Odds Ratios with 95% Confidence Intervals", x = "Odds Ratio", y = "Variable") +
  theme_minimal()
```

Physical activity: The odds of being diagnosed with COPD in individuals who did physical activity in last 30 days is 0.43 times lower than the odds of COPD in individuals who did not perform any physical activity in last 30 days (OR = 0.43, p = 0.000). This result is statistically significant.

Income category: The odds of being diagnosed with COPD in individuals whose income level is more than 50,000/year is 0.47 times lower than the odds of COPD in individuals whose income level is less than 50,000/year (OR = 0.47, p = 0.000). This result is statistically significant.

Age category: The odds of being diagnosed with COPD in individuals whose age is 65 years or older is 2.16 times higher than the odds of COPD in individuals whose whose age is less than 65 years (OR = 2.16, p = 0.000). This result is statistically significant.

Smoking status: The odds of being diagnosed with COPD in individuals who is a former smoker is 0.33 times lower than the odds of COPD in individuals who is a current smoker. (OR = 0.33, p = 0.000). This result is statistically significant.



########the end













